Analysis of Programming Students' Prompts to Identify Interactions Patterns with Generative AI

### 2024-08-28 CSEDU2025/Paper.tex
update format

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote.
% If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{array}
\usepackage{multirow}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Analysis of Programming Students' Prompts to Identify Interactions
Patterns with Generative AI \\
{\footnotesize \textsuperscript{}}
\thanks{The authors would like to thank the Federal Institute of Education,
Science, and Technology of Rio Grande do Sul (IFRS) for the partial financial
support provided for the execution of this research.}
}

\author{\IEEEauthorblockN{1\textsuperscript{st} Rodrigo Prestes Machado}
\IEEEauthorblockA{\textit{dept. Informática} \\
\textit{Instituto Federal do Rio Grande do Sul}\\
Porto Alegre, Brazil \\
0000-0003-0428-6387}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Carlos Alario Hoyos}
\IEEEauthorblockA{\textit{Dep. Ingeniería Telemática} \\
\textit{Universidad Carlos III de Madrid}\\
Madrid, Spain \\
0000-0002-3082-0814}
\and
\IEEEauthorblockN{3\textsuperscript{rd} Carlos Delgado Kloos}
\IEEEauthorblockA{\textit{Dep. Ingeniería Telemática} \\
\textit{Universidad Carlos III de Madrid}\\
Madrid, Spain \\
0000-0003-4093-3705}
}

\maketitle

\begin{abstract}
This document is a model and instructions for \LaTeX.
This and the IEEEtran.cls file define the components of your paper [title, text, heads, etc.]. *CRITICAL: Do Not Use Symbols, Special Characters, Footnotes, 
or Math in Paper Title or Abstract.
\end{abstract}

\begin{IEEEkeywords}
component, formatting, style, styling, insert
\end{IEEEkeywords}

\section{Introduction}

@@

The study by Chan and Hu et al. \cite{chan23} revealed that both undergraduate
and postgraduate students exhibit positive attitudes toward the use of GenAI
in teaching and learning, noting that it enhances the depth of their thinking
and understanding. Additionally, a systematic review by Lo, Hew, and Jong
\cite{Lo24} demonstrated that students could effectively learn from ChatGPT,
leading to improved understanding and academic achievement. Besides of that,
it was also observed  that ChatGPT enables students to regulate their learning
pace, promoting self-regulation \cite{Baha24} \cite{cai23}.

In the other hand, researchers have concerns about the impact of these
tools on students. The systematic review of Vargas-Murillo et al.
\cite{Murillo23} indicated that ChatGPT use might lead to an overreliance on the
tool. Chan et al. \cite{chan23} also noted that this reliance could result in a
decrease in critical thinking, as students might make decisions based solely on
the information provided by ChatGPT.

The student's confidence is warranted, as shown by Puryear and Sprint
\cite{Puryear22}, who demonstrated that GitHub Copilot can generate solutions
for student assignments with accuracy rates ranging from 68\% to 95\%. However,
this raises concerns that students might rely too heavily on GenAI tools,
potentially neglecting a deeper understanding of the underlying concepts.
Cai et al. \cite{cai23} further identified overdependence and reduced
intellectual engagement as significant drawbacks of using ChatGPT in learning.

Regardless of teachers' preferences or beliefs, preliminary surveys by Dickey,
Bejarano, and Garg \cite{Dickey24} revealed that at least 54.5\% of students are
already using Generative AI (GenAI) for homework. This highlights the need to
understand how students are interacting with these tools and how they can be
used to enhance learning. Furthermore, a review by Lo, Hew, and Jong \cite{Lo24}
emphasized the necessity for extended studies and objective measures to gather
more robust evidence on the use of GenAI tools in education.

@@

To deal with this issue, some researchers have proposed different pedagogical
strategies. Denny et al. \cite{Denny24} introduced the concept of \textit{Prompt
Problems}, where students solve programming exercises by formulating natural
language prompts. Prasad and Aamod \cite{Prasad24} proposed a self-regulated
learning framework using GenAI to solve programming problems. Lauren and Watta
\cite{Lauren23} explored the integration of GenAI with evidence-based
learning strategies in computer science and engineering courses.

@@

The study of effective educational strategies in use of GenAI tools in education
context are essential for determining the best training for educators. However,
as these tools gain popularity among students \cite{Dickey24} the urgency to
equip educators with best practices may lag behind their rapid adoption.
Therefore, analyzing students' interaction patterns with GenAI tools is
essential for understanding how these tools are being used and how educators can
leverage them to enhance learning. In addition, these interaction patterns can
create new opportunities to redesign GenAI tools to better support pedagogical
strategies and improve overall learning.

@@

This study aims to analyze the interactions between students and CharlieBot,
an ChatGPT-based bot developed by UC3M that leverages Retrieval-Augmented
Generation (RAG) to enhance its contextual understanding of Java programming.

\section{Related Work}

The recent advancements in Generative Artificial Intelligence (GenAI) and its
use in programming courses have raised possibilities and concerns among
educators and researchers. The study of \cite{Yilmaz23} found that the main
benefit of using ChatGPT in programming learning is to provide providing fast
and mostly correct answers to questions, improving thinking skills, facilitating
debugging, and increasing self-confidence. In contrast, the study the same
appoint limitations such ask laziness, being unable to answer some questions,
or giving incomplete/incorrect answers, causing professional anxiety in students.

The paper \cite{Lo24} From the cognitive perspective, students were able to
learn from ChatGPT, which increased their understanding and achievement.
However, concerns were raised that the growing use of AI tools might lead to
a decline in critical thinking among students. Teachers and stakeholders should
continue to investigate pedagogical approaches that leverage ChatGPT to enhance
students' understanding and critical thinking. For example, teachers can
instruct students to fact-check and validate information produced by ChatGPT.

examined students from different demographic groups
using AI tools like ChatGPT, comparing perceptions and impacts
among these groups. The results showed that AI tools had a positive impact on
academic performance, especially for students with less prior experience.

@@

\subsection{The CharlieBot}

@@

\begin{table*}[htbp]
\caption{Categories of Analysis}
    \begin{center}
    \begin{tabular}{|p{3cm}|p{5cm}|p{3cm}|}
    \hline
    \textbf{Category} & \textbf{Description} & \textbf{Critical Thinking} \\
    \hline
    Debugging Help & Prompts that seek help to identify, fix errors, or understand the provided code snippet. & \multirow{6}{3cm}{Tends to promote.} \\
    Conceptual Questions & Prompts that are more about understanding concepts or algorithms rather than specific code. & \\
    Student Correction & Prompts where the student corrects the bot. & \\
    \hline
    Code Snippet & Prompts that ask for a specific part of the code, like a function or a segment. & \multirow{6}{3cm}{Tends not to promote.} \\
    Complete Solution & Prompts that request an entire solution or a complete code snippet. & \\
    Multiple Question & Prompts where the user wants to solve a multiple-question exercise. & \\
    \hline
    Language Change & Prompts that request a change of idiom to convey a message more effectively. & \multirow{3}{3cm}{Neutral.} \\
    Uncategorized & Prompts that do not fit into any of the above categories. & \\
    \hline
    \end{tabular}
    \label{tab:categories}
    \end{center}
\end{table*}

\section{Method}

@@

@@

The graph \ref{fig:graph1} shows the distribution of the messages in the
categories:
\begin{figure}[h!]
    \centering
    \includegraphics[scale=0.7]{figures/figure1.png}
    \caption{Classification of messages.}
    \label{fig:graph1}
\end{figure}

@@

\bibliographystyle{IEEEtran}
\bibliography{references}

>>>>

\documentclass[a4paper,twoside]{article}

\usepackage{epsfig}
\usepackage{subcaption}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{multicol}
\usepackage{pslatex}
\usepackage{algorithm2e}
\usepackage[bottom]{footmisc}
% Please add other packages that you may need BEFORE the SCITEPRESS.sty package.
\usepackage{natbib}
\usepackage{multirow}
\usepackage{SCITEPRESS}

\begin{document}

\title{Analysis of Programming Students' Prompts to Identify Interactions
Patterns with Generative AI}

\author{\authorname{Rodrigo Prestes Machado\sup{1}\orcidAuthor{0000-0003-0428-6387},
Carlos Alario Hoyos\sup{2}\orcidAuthor{0000-0002-3082-0814} and
Carlos Delgado Kloos\sup{2}\orcidAuthor{0000-0003-4093-3705}}
\affiliation{\sup{1}Department of Informatics, Federal Institute of Education,
Science and Technology, Porto Alegre, Brazil}
\affiliation{\sup{2}Department Telematics Engineering, Carlos III University,
Madrid, Spain}
\email{\ rodrigo.prestes@poa.ifrs.edu.br, \{calario,cdk\}@it.uc3m.es}
}

\keywords{Generative Artificial Intelligence, Programming Education, Chatbots}

\abstract{The abstract should summarize the contents of the paper and should
contain at least 70 and at most 200 words. The text must be set to 9-point font
size.}

\twocolumn\maketitle\normalsize\setcounter{footnote}{0}

\section{\uppercase{Introduction}}
\label{sec:introduction}

@@

The study of \cite{chan23} revealed that both undergraduate and postgraduate
students exhibit positive attitudes toward the use of GenAI in teaching and
learning, noting that it enhances the depth of their thinking and understanding.
Additionally, a systematic review conducted by \cite{Lo24} demonstrated that
students could effectively learn from ChatGPT, leading to improved understanding
and academic achievement. Besides of that, it was also observed that ChatGPT
enables students to regulate their learning pace, promoting self-regulation
\citep{Baha24} \citep{cai23}.

In the other hand, researchers have concerns about the impact of these
tools on students. The systematic review of \cite{Murillo23} indicated that
ChatGPT use might lead to an overreliance on the tool. \cite{chan23} also noted
that this reliance could result in a decrease in critical thinking, as students
might make decisions based solely on the information provided by ChatGPT.

The student's confidence is warranted, as shown by \cite{Puryear22}, who
demonstrated that GitHub Copilot can generate solutions for student assignments
with accuracy rates ranging from 68\% to 95\%. However, this raises concerns that
students might rely too heavily on GenAI tools, potentially neglecting a deeper
understanding of the underlying concepts. \cite{cai23} further identified
overdependence and reduced intellectual engagement as significant drawbacks of
using ChatGPT in learning.

Regardless of teachers' preferences or beliefs, preliminary surveys
performed by \cite{Dickey24} revealed that at least 54.5\% of students are
already using GenAI for homework. This highlights the need to understand how
students are interacting with these tools and how they can be used to enhance
learning. Furthermore, a review by \cite{Lo24} emphasized the necessity for
extended studies and objective measures to gather more robust evidence on the
use of GenAI tools in education.

@@

To deal with this issue, some researchers have proposed different pedagogical
strategies \cite{Denny24} introduced the concept of \textit{Prompt
Problems}, where students solve programming exercises by formulating natural
language prompts. \cite{Prasad24} proposed a self-regulated learning framework
using GenAI to solve programming problems. \cite{Lauren23} explored the
integration of GenAI with evidence-based learning strategies in computer science
and engineering courses.

@@

The study of effective educational strategies in use of GenAI tools in education
context are important for determining the best training for educators. However,
as these tools gain popularity among students \cite{Dickey24} the urgency to
equip educators with best practices may lag behind their rapid adoption.
Therefore, analyzing students' interaction patterns with GenAI tools is
important for understanding how these tools are being used and how educators can
leverage them to enhance learning. In addition, these interaction patterns can
create new opportunities to redesign GenAI tools to better support pedagogical
strategies without repressing the development of abstract, critical and creative
thinking.

@@

This study aims to analyze the interactions between students and CharlieBot,
an ChatGPT 3.5 based bot developed by UC3M that leverages Retrieval-Augmented
Generation (RAG) to enhance its contextual understanding of Java programming.

\section{\uppercase{Related Work}}

The advancements in GenAI and its use in programming courses have raised
possibilities and concerns among educators and researchers. The study of
\cite{Yilmaz23} found that the main benefit of using ChatGPT in programming
learning is to provide providing fast and mostly correct answers to questions,
improving thinking skills, facilitating debugging, and increasing
self-confidence. In contrast, the study the same appoint limitations such ask
laziness, being unable to answer some questions, or giving incomplete/incorrect
answers, causing professional anxiety in students.

From a cognitive perspective, \cite{Lo24} shown that students were able to
learn from ChatGPT, which increased their understanding and achievement.
However, concerns were raised that the growing use of AI tools might lead to a
decline in critical thinking among students. Teachers and stakeholders should
continue to investigate pedagogical approaches that leverage ChatGPT to enhance
students' understanding and critical thinking. For example, teachers can
instruct students to fact-check and validate information produced by ChatGPT.

Examined students from different demographic groups using AI tools like
ChatGPT, comparing perceptions and impacts among these groups. The results
showed that AI tools had a positive impact on academic performance, especially
for students with less prior experience.

@@

\subsection{CharlieBot}

@@

\section{\uppercase{Method}}

@@

\begin{table*}[htbp]
  \caption{Categories of Analysis}
      \begin{center}
      \begin{tabular}{|p{4cm}|p{6cm}|p{3cm}|}
      \hline
      \textbf{Category} & \textbf{Description} & \textbf{Critical Thinking} \\
      \hline
      Debugging Help & Prompts that seek help to identify, fix errors, or understand the provided code snippet. & \multirow{7}{3cm}{Tends to promote.} \\
      Conceptual Questions & Prompts that are more about understanding concepts or algorithms rather than specific code. & \\
      Student Correction & Prompts where the student corrects the bot. & \\
      \hline
      Code Snippet & Prompts that ask for a specific part of the code, like a function or a segment. & \multirow{6}{3cm}{Tends not to promote.} \\
      Complete Solution & Prompts that request an entire solution or a complete code snippet. & \\
      Multiple Question & Prompts where the user wants to solve a multiple-question exercise. & \\
      \hline
      Language Change & Prompts that request a change of idiom to convey a message more effectively. & \multirow{4}{3cm}{Neutral.} \\
      Uncategorized & Prompts that do not fit into any of the above categories. & \\
      \hline
      \end{tabular}
      \label{tab:categories}
      \end{center}
  \end{table*}

@@

The Figure \ref{fig:graph1} shows the distribution of the messages in the
categories:

\begin{figure}[h!]
    \centering
    \includegraphics[scale=0.55]{img/figure1.png}
    \caption{Classification of messages.}
    \label{fig:graph1}
\end{figure}

@@

\section*{ACKNOWLEDGEMENTS}

The authors would like to thank the Federal Institute of Education, Science and
Technology (IFRS) for the partial financial support provided for the execution
of this research.

\bibliographystyle{apalike}
{\small
\bibliography{References}}

\section*{\uppercase{Appendix}}

If any, the appendix should appear directly after the
references without numbering, and not on a new page. To do so please use the
following command: \textit{$\backslash$section*\{APPENDIX\}}
